---
title: "Data wrangling: one data table" 
author: "ECON 122"
date: "Day 6"
output: 
  ioslides_presentation:
    incremental: true
    widescreen: true
    fig_height: 4
    fig_width: 7
    keep_md: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, prompt=TRUE, eval=TRUE, message=F, include=T,comment=NULL)
```

```{r packageCheck, include=FALSE}
mypacks <- c("ggplot2","dplyr", "lubridate")  # what packages are needed?
packs <- installed.packages()   # find installed package list
install.me <- mypacks[!(mypacks %in% packs[,"Package"])]  #what needs to be installed?
if (length(install.me) >= 1) install.packages(install.me, repos = "http://cran.us.r-project.org")   # install (if needed)
lapply(mypacks, library, character.only=TRUE)  # load all packages
```

## Data Wrangling {.build}

- Data wrangling is an extremely important data science skill
- methods for *filtering*, *transforming*,  *restructuring* and *combining* datasets should be 
    + readable
    + reproducible
    + include checks of data quality (typos, code errors, etc)
- Hadley Wickham's `dplyr` package is designed for this type of wrangling

## The `dplyr` package {.build}

- package is designed for readable data wrangling
    + everything you can do with `dplyr` can be done without it (but maybe not as nicely/easily)
- There are 5 key *verbs* (actions):
    + `filter()`: pick observations (rows) of interest (subset)
    + `select()`: pick variables (columns)
    + `mutate()`: modify existing variables or create new variables
    + `summarize()`: collapse many values down into a single summary 
    + `arrange()`: sort obervations
- `group_by()`: changes the scope, or unit of analysis, from individual cases to aggregated groups
    - use with either `summarize` or `mutate`

## `filter` {.build}

- Basic command: `filter(data, condition)` 
- `filter(data, cond1, cond2)`: rows that meet `cond1` AND `cond2`
- Let's filter (subset) the `loans` data to get the 300 bad loans cases:
```{r}
loans <- read.csv("https://raw.githubusercontent.com/mgelman/data/master/CreditData.csv")
badLoans  <- filter(loans, Good.Loan == "BadLoan")
dim(badLoans)
badLoans2 <- loans[loans$Good.Loan == "BadLoan",]
dim(badLoans2)
```

- What about bad loans that are costly?!
```{r}
costlyBadLoans <- filter(loans, Good.Loan == "BadLoan", Credit.amount > 10000)
dim(costlyBadLoans)
```

## `select` {.build}

- Basic command: 
    - `select(data, var1, var2)` gets vars 1 and 2 (add `-` to omit)
    - `select(data, var3:var5)` gets vars 3-5 (add `-` to omit)
    - `select(data, starts_with("ab"))` select variables starting with "ab"
    - `select(data, ends_with("yz"))` select variables ending with "yz"
    - `select(data, contains("mn"))` selects variables containing "mn" strings
    
- Omit variables with a period . in the name:
```{r}
loansNoPeriods <- select(loans, -contains("."))
names(loansNoPeriods)
```
    
    
## `select` {.build}

- Get credit amount from `loans`
```{r}
credit <- select(loans, Credit.amount)
str(credit)  # one vector is still a data frame! 
```
-  `dplyr` commands preserve data frame class
- base-R subsetting does not always preserve class
```{r}
credit2 <- loans[, "Credit.amount"]
str(credit2)  
```

## Piping  {.build}

- piping `%>%` allows you to put the output of one function into the input of another
    - akin to function composition
    - `x %>% f() %>% g()` is equal to `g(f(x))`
    - `x %>% f() %>% g()` is read: take `x`, then apply function `f`, then apply function `g`
    - Think of it like an assembly line where you pass the completed task on to another worker
    
## Piping  {.build}

- Let's pull the credit and duration for bad loans
```{r}
badloans <- loans %>% 
  filter(Good.Loan == "BadLoan") %>%
  select(Credit.amount, Duration.in.month)
```
is the same as
```{r}
badLoans2  <- select(filter(loans, Good.Loan == "BadLoan"), Credit.amount, Duration.in.month)
```

- Same but more readable (means less likely to make errors)


## `mutate` {.build}

- Basic command: `mutate(data, newvar1= fun1, newvar2= fun2)`
- Adds a newly created variable(s) to `data`
    + can use `newvar1` to create `newvar2` in the same command
- Called a *vectorized* command because it applies a function to all individual variable entries (preserved data frame length)
```{r}
loans %>%
            mutate(durationYears = Duration.in.month/12) %>%
            select(Duration.in.month, durationYears)
```

## `mutate` {.build}

- base-R: construct loan default predictions
```{r}
loans$pred.Def1 <- ifelse(loans$Duration.in.month > 24 & loans$Credit.amount > 10000,
                          "predBad", "predGood")
loans$pred.Def <- ifelse(loans$Duration.in.month <= 24 & loans$Credit.amount < 2200,
                         "predBad", loans$pred.Def1)
head(loans[, c("Duration.in.month","Credit.amount","pred.Def1","pred.Def","Good.Loan")], 6)
```
How can we do this using `dplyr`?

## `mutate` {.build}

- `dplyr`: construct loan default predictions
```{r}
loans2 <- loans %>%
            mutate(pred.Def1 = ifelse(Duration.in.month > 24 & Credit.amount >10000,
                                      "predBad", "predGood"), 
                   pred.Def = ifelse(Duration.in.month <= 24 & Credit.amount < 2200,
                                     "predBad", pred.Def1)) %>%
            select(Duration.in.month, Credit.amount, pred.Def1, pred.Def, Good.Loan)
head(loans2, 10)
```


## `summarize` {.build}

- Basic command: `summarize(data, newvar1= fun1(var), newvar2= fun2(var))`
- `summarize` applies the `fun` to variable(s) and returns *one* value
    - takes a data frame with `n` rows and returns a data frame with 1 row
- Here are the mean and sd for credit amount
```{r}
loans %>% summarize(mn = mean(Credit.amount), sd = sd(Credit.amount))
```

## `summarize` {.build}

- proportion of *all loans that defaulted* that are less or equal to 24 months duration 
- `base-R`
```{r}
mean(loans$Duration.in.month[loans$Good.Loan == "BadLoan"] <= 24)
```
- `dplyr`
```{r}
loans %>% filter(Good.Loan == "BadLoan") %>%
  mutate(less24 = Duration.in.month <= 24) %>%
  summarize(prop = mean(less24))
```

## `group_by`  {.build}

- the most awesome `dplyr` feature is the `group_by` command
- `group_by(groupVar1, groupVar2)` changes the analysis unit from the individuals to the groups defined by `groupVar`s
    - followed by `summarize`, gives stats by groups
    - followed by `mutate`, applies function by group
- can `ungroup()` to return "focus" back to entire data frame (instead of groups)


## `group_by` {.build}

-  credit amounts by loan status
    - `n()` gives us the number of observations
```{r}
loans.summarize <- loans %>% group_by(Good.Loan) %>%
             summarize(mn = mean(Credit.amount), sd = sd(Credit.amount), n = n())
loans.summarize
```


## Tibbles {.build}

- Hadley has also created a new type of "opinionated" data frame called a **tibble**
```{r}
class(loans.summarize)
glimpse(loans.summarize)
```
- the `group_by` command will always convert a base-R `data.frame` to a tibble data frame

## Tibbles {.build}

- why use tibbles?
    + smart view by calling data frame (only shows a data snapshot rather than entire data frame)
    + preserves object type, subsetting always yields another tibble
```{r}
loans.summarize[1,1]
```

## `group_by` {.build}

- can include more than one grouping variable
```{r}
loans2 %>% group_by(Good.Loan, pred.Def) %>% print(n=2)
```

## `group_by` {.build}

- Accuracy rate = % of all loans that were correctly predicted to default or not
- How can we use `group_by` to calculate the accuracy rate?
```{r}
loans2 %>% group_by(Good.Loan, pred.Def) %>%
  summarize(n = n(), prop = n/nrow(.))
```
- The period `.` inserts the data frame used in that step
- Accuracy rate: 13.7% + 34.5% = `r 13.7 + 34.5`%

## `group_by` {.build .smaller} 

- What if we want false positive (default) and false negative rate? 
  - false positive rate: % of good loans that were predicted to default
  - false negative rate: % of bad loans that were predicted to not default
- Instead of dividing by overall `n` we need to divide by `n` within `Good.Loan`
- additional `group_by` will replace previous groupings
```{r} 
loans2 %>% group_by(Good.Loan, pred.Def) %>%  summarize(n = n()) %>%
  group_by(Good.Loan) %>%
  mutate(condProp = n/sum(n))
```
- false positive rate: 50.7% of good loans were predicted to default
- false negative rate: 54.3% of bad loans were predicted to not default


## `arrange` {.build}
- How can we sort values in a DF?
- One last command: `arrange(data, var1)` sorts from low to high by `var1`
    - descending order uses `desc(var1)`
```{r}
loans %>% select(Credit.amount, Good.Loan,Duration.in.month) %>%
  arrange(desc(Credit.amount)) %>% head()
```


